Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-27 Manuscript/manuscript.Rmd
Clean up ??? in Results

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. After removing TADs that exceeded 2 md in length (biologically plausible maximum TAD size [@dixon2012topological; @sofueva2013cohesin] [??? Add ref, see one in SpectralTAD; !!!added]), we obtained a total of 16133, 14824, 8081, 3560, and 919 unique TAD boundaries for the GM12878 cell line, and 9314, 10881, 6923, 3197, and 831 unique TAD boundaries for the K562 cell line across each of the five resolutions respectively (Supplementary Figure 1). To account for the variability in boundary location due to varying resolution, we developed a resolution-specific approach toward constructing TAD versus non-TAD boundary regions on the linear genome (See Methods), labeled as a response vector **Y** (Figure 1A). TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$). On average, we found that only 18.0% (SD=4.6%) [??? Need to put exactly, e.g., $18.34 \pm 5.06 \%$; !!!added] of TAD boundary regions overlapped between both cell lines (Supplementary Figure 2). Given the cell type specificity of TAD boundaries, we opted to build cell line specific models for predicting TAD boundary regions. The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material. In summary, we developed a framework enabling resolution-specific prediction of TAD boundaries, with resolution-specific bins defining the response variable.

@@

Given that the number of bins annotated as TAD boundaries (TAD boundary regions) is much smaller than the number of non-TAD boundary bins, we investigated the class imbalance (CI) [??? I changed PI to CI - change in supplementary accordingly; !!!fixed] featured at each resolution. The CI was defined as the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions and ranging from 0 to 0.50. The CI was found to be invariant to resolution, remaining consistently low across chromosomes, with an average of 0.196 (SD=0.025) and 0.166 (SD=0.026) for the GM12878 and K562 cell lines, respectively [??? Add exact numbers, like $18.34 \pm 5.06 \%$ !!!added] (Supplementary Figure 3). To account for such imbalance we evaluated 4 re-sampling techniques (in addition to no re-sampling) before predicting TAD boundaries. **Random over-sampling** (ROS) was defined as sampling with replacement from TAD boundary regions. **Random under-sampling** (RUS) was defined as sampling with replacement from non-TAD boundary regions. Lastly, we considered **Synthetic minority over-sampling technique** (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote].

[??? Very briefly introduce each technique, refer to Methods for details; !!!added]

@@

To identify the combination of data resolution, predictor type, and re-sampling technique necessary for optimal cell line specific TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 2). The framework uses data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, using one of the predictor type *P = {OC, OP, Distance}*, and the re-sampling technique *B = {None, RUS, ROS, SMOTE}*. Once the model inputs were established, a random forest classifier was trained on chromosome-specific datasets using 10-fold cross validation. The different combinations of inputs were evaluated by assessing the Matthews Correlation Coefficient (MCC) of each model [??? Ugly sentence, rewrite; !!!fixed]. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2 \times 5 \times 3 \times 4 \times 21 = 2520$) models. Additional performance metrics are reported in the Supplementary materials.

![**A diagram of the ensemble framework detailing the combination of inputs used in the model building pipeline.** Each cell line (GM12878, K562) specific model was composed of a combination of different data level characteristics including resolution (5 kb, 10 kb, 25 kb, 50 kb, 100 kb), predictor type (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)), and re-sampling technique (no re-sampling (None), random over-sampling (ROS), random under-sampling (RUS), synthetic minority over-sampling technique (SMOTE)). The following sets of genomic annotations were used to create each predictor type including histone modification (HM), chromating state segmentations (ChromHMM), transcription factor binding sites (TFBS). [??? Describe abbreviations on the figure, the figure itself is already described in text;!!!fixed] ](Figures/figure3_ensemble_framework_V3.png)

## SMOTE RESAMPLING, DISTANCE-BASED ASSOCIATIONS OF TAD BOUNDARIES AND GENOMIC ANNOTATIONS, AND 10KB RESOLUTION OF HI-C DATA PROVIDED OPTIMAL MODEL PERFORMANCE [??? This paragraph was completely rewritten, check for logic flow; !!!checked]

@@

Interestingly, we found that performances peaked at 10 kb resolution and then degraded as data resolution decreased. The reason for the improved performance between 5 kb and 10 kb data resolutions can be due to the sparsity in the interaction frequency matrices at finer resolutions. [??? Quantify sparsity, need to substantiate this statement; !!!divide nonzero elements by total?] This sparsity may lead to inaccurate TAD calling and, subsequently, to worse model performance. These results suggest that finer resolution of Hi-C data is not always optimal for accurate TAD calling and prediction, and sparsity of the data needs to be considered. Additionally, performances tended to be much more variable between chromosomes at lower resolutions, evidenced by the wider error bars for the 50 kb and 100 kb resolutions. [??? Where are the results?; !!!added]

In summary, our results imply that using random under- and over-sampling, distance-type predictors, and the data at 10kb resolution results in an optimally performing cell line-specific random forest-based TAD boundary prediction model.

@@

For each predicted TAD boundary region, we were able to identify the summit locations of TAD boundary location by performing partitioning around medoids (see Methods). To compare the summit locations of predicted and called TAD boundaries, we evaluated the signal strength of ChIP-seq peak regions around a 10 kb flanking region. We found that the signal strength for the CTCF, RAD21, SMC3, and ZNF143 [??? Make all in this order!; !!!reasoning of order is to match heatmap; discuss in meeting] TFBS peaked at our predicted TAD boudaries, while was poorly defined around the called TAD boundaries (Figure 5; Supplementary Figure 10[???; !!!fixed]).  Thus, our model is able to alleviate resolution limitations and precisely locate TAD boundaries that correspond to the highest signal strength of known TAD boundary-forming TFBSs.

>>>>

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. After removing TADs that exceeded 2 md in length (biologically plausible maximum TAD size [@dixon2012topological; @sofueva2013cohesin]), we obtained a total of 16133, 14824, 8081, 3560, and 919 unique TAD boundaries for the GM12878 cell line, and 9314, 10881, 6923, 3197, and 831 unique TAD boundaries for the K562 cell line across each of the five resolutions respectively (Supplementary Figure 1). To account for the variability in boundary location due to varying resolution, we developed a resolution-specific approach toward constructing TAD versus non-TAD boundary regions on the linear genome (See Methods), labeled as a response vector **Y** (Figure 1A). TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$). On average, we found that only $18.0 \pm 4.6 \%$ of TAD boundary regions overlapped between both cell lines (Supplementary Figure 2). Given the cell type specificity of TAD boundaries, we opted to build cell line specific models for predicting TAD boundary regions. The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material. In summary, we developed a framework enabling resolution-specific prediction of TAD boundaries, with resolution-specific bins defining the response variable.

@@

Given that the number of bins annotated as TAD boundaries (TAD boundary regions) is much smaller than the number of non-TAD boundary bins, we investigated the class imbalance (CI) featured at each resolution. The CI was defined as the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions and ranging from 0 to 0.50. The CI was found to be invariant to resolution, remaining consistently low across chromosomes, with an average of $0.196 \pm 0.025$ and $0.166 \pm 0.026$ for the GM12878 and K562 cell lines, respectively (Supplementary Figure 3). To account for such imbalance, we evaluated three re-sampling techniques (in addition to no re-sampling) before predicting TAD boundaries. **Random over-sampling** (ROS) was defined as sampling with replacement from TAD boundary regions. **Random under-sampling** (RUS) was defined as sampling with replacement from non-TAD boundary regions. Lastly, we considered **Synthetic minority over-sampling technique** (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote].

@@

To identify the combination of data resolution, predictor type, and re-sampling technique necessary for optimal cell line specific TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 2). The framework uses data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, using one of the predictor type *P = {OC, OP, Distance}*, and the re-sampling technique *B = {None, RUS, ROS, SMOTE}*. Once the model inputs were established, a random forest classifier was trained on chromosome-specific datasets using 10-fold cross validation. The different combinations of inputs were evaluated by assessing the Matthews Correlation Coefficient (MCC) of each model. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2 \times 5 \times 3 \times 4 \times 21 = 2520$) models. Additional performance metrics are reported in the Supplementary materials.

![**A diagram of the ensemble framework detailing the combination of inputs used in the model building pipeline.** Each cell line (GM12878, K562) specific model was composed of a combination of different data level characteristics including resolution (5 kb, 10 kb, 25 kb, 50 kb, 100 kb), predictor type (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)), and re-sampling technique (no re-sampling (None), random over-sampling (ROS), random under-sampling (RUS), synthetic minority over-sampling technique (SMOTE)). The three types of genomic annotations, histone modification (HM), chromating state segmentations (ChromHMM), transcription factor binding sites (TFBS), were used as predictors.](Figures/figure3_ensemble_framework_V3.png)

## SMOTE RESAMPLING, DISTANCE-BASED ASSOCIATIONS OF TAD BOUNDARIES AND GENOMIC ANNOTATIONS, AND 10KB RESOLUTION OF HI-C DATA PROVIDED OPTIMAL MODEL PERFORMANCE

@@

Interestingly, we found that performances peaked at 10 kb resolution and then degraded as data resolution decreased. The reason for the improved performance between 5 kb and 10 kb data resolutions can be due to the sparsity in the interaction frequency matrices at finer resolutions. [??? Quantify sparsity, need to substantiate this statement; !!!divide nonzero elements by total?] This sparsity may lead to inaccurate TAD calling and, subsequently, to worse model performance. These results suggest that finer resolution of Hi-C data is not always optimal for accurate TAD calling and prediction, and sparsity of the data needs to be considered. Additionally, performances tended to be much more variable between chromosomes at lower resolutions, evidenced by the wider error bars for the 50 kb and 100 kb resolutions. In summary, our results imply that using random under- and over-sampling, distance-type predictors, and the data at 10kb resolution results in an optimally performing cell line-specific random forest-based TAD boundary prediction model.

@@

For each predicted TAD boundary region, we were able to identify the summit locations of TAD boundary location by performing partitioning around medoids (see Methods). To compare the summit locations of predicted and called TAD boundaries, we evaluated the signal strength of ChIP-seq peak regions around a 10 kb flanking region. We found that the signal strength for the CTCF, RAD21, SMC3, and ZNF143 [??? Make all in this order!; !!!reasoning of order is to match heatmap; discuss in meeting] TFBS peaked at our predicted TAD boudaries, while was poorly defined around the called TAD boundaries (Figure 5; Supplementary Figure 10).  Thus, our model is able to alleviate resolution limitations and precisely locate TAD boundaries that correspond to the highest signal strength of known TAD boundary-forming TFBSs.
